### 2020-08-29 thesis/thesis.tex
[thesis] Various corrections and improvements

In this thesis we discuss issues arising from combining algebraic effects
and handlers with resource management, and present a solution through coalgebraic
means along with presentation of experimental programming language Freak, which
is a language with (co)algebraic effects and (co)handlers, where implementation
is based on Continuation Passing Style translation.

@@

They give the developer necessary tools for declarative programming with
computational effects on a high level of abstraction, by defining
effects as an API\@ in the code. At the same time, algebraic effects preserve
a lot of flexibility and have a strong theoretical foundations.

@@

In the following chapter we provide a background about effects, define algebraic
effects in the categorical setting as well as point out dual coalgebraic effects
and cohandlers, finishing with showing related work in this area.
Chapter~\ref{chapter:potential-solutions} describes possible ways to approach
the issue of excessive generality, Chapter~\ref{chapter:co-effectful-programming}
shows the Freak language by examples along with a usage guide, and then
Chapter~\ref{chapter:calculus-of-freak-language} describes the language's syntax,
operational semantics and CPS translation. In the next one,
Chapter~\ref{chapter:implementation}, implementation details are revealed.
We conclude in Chapter~\ref{chapter:conclusion} by stating what the possible
augmentations are, which are intended to be made in the future.

@@

It turns out, that it becomes complex and unpleasant to combine two functors
together to form a new monad, and for this purpose, monad
transformers~\cite{monad-transformers} arose in Haskell. Most of the common
languages avoid this by not expressing computational effects in the type system,
and instead one may think about functions as being implicitly embed in a Kleisli
Category over a functor T~\cite{kleisli-ncatlab}, where T is a hidden signature
over all possible side effects that occur in our program.

@@

That being said, it's only one particular issue that algebraic effects address.
Their handlers may also drop the resumption or invoke it more than
once. There are many other great sources for getting familiarized with
effects from practical side~\cite{handlers-tutorial, programming-in-eff, koka-tutorial},
so we will omit further explanations. More examples can be found in
Chapter~\ref{chapter:co-effectful-programming}.

@@

@@

        A $ \mathVar{\Sigma-Equation} $ is an object X and a pair of
        $\mathVar{\Sigma-terms}$ $l, r \in Tree_{\Sigma}(X)$, written as

@@

    A \textit{model M} of an algebraic theory T is an interpretation of the
    signature ${\Sigma_{T}}$ which validates all the equations $\mathcal{E_{T}}$.
    That is, for every equation $X \mid l = r$ the following diagram commutes:

@@

    Let L, M be models of a theory T. $ \mathVar{T-homomorphism}$
    $\varphi : L \rightarrow M$ is a map such that, for every operation symbol
    $op_{i}$ in T the following diagram commutes:

@@

    Free F-algebra on an object A (of generators) in $\mathcal{C}$ is meant an algebra

@@

    \begin{definition}
        Comodel in $\mathcal{C}$ is just a Model in $ \mathcal{C}^{op} $.
    \end{definition}

@@

    % TODO: textbf Set^op
    \noindent
    We want now to embed this morphism in $Set^{op}$. Every arrow in opposite
    category is reverted, thus, products become coproducts and our morphism
    is reverted:

@@

    We have now derived a dual concept to algebraic effects and handlers, which
    in the literature are also called comodels and runners~\cite{runners-uustalu, runners-in-action}.
    As we have seen, comodels turned into state-passing (co)operations, which
    in programming languages, may be a way to model interaction with external
    resources, where based on current configuration and some parameter, we
    obtain new configuration along with it's result.

@@

    Induction is a way of constructing new structures. Recursion is a way of
    folding inductively defined structure in a terminating way. Recursive
    functions (not to be confused with recursive computability class), should
    shrink the argument in each call, meaning that it eventually ends up
    terminating in a base case.

    Coinduction is literally a dual notion to induction. We \textit{observe} possibly
    infinite structures, by doing deconstruction. Corecursion, is a way
    of productively defining new, possibly enlarged structures. Due to infinity,
    the evaluation should be lazy, whereas in induction it may be eager.

@@

\noindent
TODO should this section be further extended?

@@

    \subsection{Cohandlers}\label{subsec:calc-cohandlers}

    Semantics for cohandlers is brought from denotational one that we have
    introduced in Section~\ref{sec:duality}, meaning, we denote cooperations
    as state passing transitions in a code, represented similarly as algebraic
    effects, through (co)free (co)models of algebraic theories along with their
    morphisms. Let's put the following cooperation:

@@

    Operationally, we embed transitions between coeffectull computations
    in a standard state monad $ S_{C}X \triangleq (C \Rightarrow X \times C) $.
    We achieve this by transforming AST of the source code, such that
    configurations are passed among cooperations. For a given coalgebraic theory
    $T$, handler of the state monad in Freak looks as follows:

@@

    \noindent
    into the following one (modulo $\mathVar{alpha-conversion}$):

\vbox{\begin{verbatim}
    let Tconf <- observe TGet () in
    let Tcoresult <- observe Coeffect (Tconf, p) in
    let TnewConf <- return fst Tcoresult in
    let Tres <- return snd Tcoresult in
    let _ <- observe TPut TnewConf in
    return Tres
\end{verbatim}
}

    \noindent
    where before executing a coeffect, we obtain current configuration,
    pass it to coeffect, put new config and then return the actual result.
    This is done behind the scenes and user does not know about state
    passing.

@@

    TODO Describe the transformation here.

@@

\begin{verbatim}
    AST.hs          - AST data structures
    CommonCPS.hs    - Common functions for CPS translation
    CommonEval.hs   - Common functions for evaluation
    CPSLists.hs     - Uncurried CPS translation
    EvalTarget.hs   - Evaluation of the target calculus
    Freak.hs        - API for the language
    Main.hs         - Main module running evaluator on given filename
    Parser.hs       - Parser and lexer
    TargetAST.hs    - AST for the target calculus
    Tests.hs        - Tests module
    Types.hs        - Common types definition
    Transform.hs    - Cohandlers transformation described in Calculus chapter
    programs/       - Exemplary programs covered in tests

@@

    We have showed derivation of cooperations, based on previous work, as well
    as presented a calculus combining algebraic effects with their dual notion.
    This resulted in a programming language which allows user to seemlessly
    operate on resources based on solid mathematical foundations. It can seen now,
    that (co)algebraic effects are an abstract concept that captures plenty of
    patterns in everyday programming, which already had ad-hoc and more
    restricted implementations in many existing languages, through constructs
    such as \verb!try, except, finally!, or by more low-level mechanisms
    such as handling interrupts in operating systems.

    This relatively new concept in theory of progamming languages yet needs to
    be researched, and we propose in the next section a few directions into
    which Freak language could go.

>>>>

In this thesis we discuss issues arising from combining algebraic effects
and handlers with resource management, and present a solution through coalgebraic
means along with presentation of Freak, experimental programming language with
(co)algebraic effects and (co)handlers. Implementation is based on Continuation
Passing Style translation.

@@

They give the developer necessary tools for declarative programming with
computational effects, on a high level of abstraction, by defining
effects as an API\@ in the code. At the same time, algebraic effects preserve
a lot of flexibility and have a strong theoretical foundations.

@@

In the following chapter we provide a background about effects, define algebraic
effects in the categorical setting, as well as point out dual coalgebraic effects
and cohandlers, finishing with showing related work in this area.
Chapter~\ref{chapter:potential-solutions} describes possible ways to approach
the issue of excessive generality, Chapter~\ref{chapter:co-effectful-programming}
shows the Freak language through examples along with a usage guide, and then
Chapter~\ref{chapter:calculus-of-freak-language} describes more formally the
language's syntax, operational semantics and CPS translation. In the next one,
Chapter~\ref{chapter:implementation}, implementation details are revealed.
We conclude in Chapter~\ref{chapter:conclusion} by stating what the possible
augmentations are, which are intended to be made in the future.

@@

It turns out, that it becomes complex and unpleasant to combine two functors
together to form a new monad, and for this purpose, monad
transformers~\cite{monad-transformers} arose in Haskell. Most of the common
languages avoid this by not expressing computational effects in the type system,
and instead one may think about functions as being implicitly embed in a Kleisli
Category over a functor $T$~\cite{kleisli-ncatlab}, where $T$ is a hidden signature
over all possible side effects that occur in our program.

@@

That being said, it's only one particular issue that algebraic effects address.
There are many other great sources for getting familiarized with effects from
practical side~\cite{handlers-tutorial, programming-in-eff, koka-tutorial},
so we will omit further explanations. More examples can be found in
Chapter~\ref{chapter:co-effectful-programming}.

@@

    $\mathVar{Free}$ $\mathVar{F-algebra}$ on an object $A$ (of generators) in $\mathcal{C}$ is meant an algebra

@@

        A $ \mathVar{\Sigma-Equation} $ is an object $X$ and a pair of
        $\mathVar{\Sigma-terms}$ $l, r \in Tree_{\Sigma}(X)$, written as

@@

    A \textit{model M} of an algebraic theory $T$ is an interpretation of the
    signature ${\Sigma_{T}}$ which validates all the equations $\mathcal{E_{T}}$.
    That is, for every equation $X \mid l = r$ the following diagram commutes:

@@

    Let $L$, $M$ be models of a theory $T$. $ \mathVar{T-homomorphism}$
    $\varphi : L \rightarrow M$ is a map such that, for every operation symbol
    $op_{i}$ in $T$ the following diagram commutes:

@@

@@

    \begin{definition}
        $\mathVar{Comodel}$ in $\mathcal{C}$ is just a Model in $ \mathcal{C}^{op} $.
    \end{definition}

@@

    \noindent
    We want now to embed this morphism in $\mathbf{Set^{op}}$. Every arrow in opposite
    category is reverted, thus, products become coproducts and our morphism
    is reverted:

@@

    We have now derived a dual concept to algebraic effects and handlers, which
    in the literature are also called comodels and runners~\cite{runners-uustalu, runners-in-action}.
    As we have seen, comodels turned into state-passing (co)operations.
    In programming languages, they may be a way to model interaction with external
    resources, where based on current configuration and a parameter, we
    obtain new configuration along with it's result.

@@

    \begin{definition}
        \textit{Initial algebra} is an initial object in \textbf{Alg}, the category of algebras
        and homomorphisms.
    \end{definition}

    \begin{lemma}
        Free algebra over an initial object is initial algebra. Dually,
        cofree coalgebra over a final object is final coalgebra.
    \end{lemma}
    \begin{proof}
        Follows immediately after putting initial object as generator in the
        diagram for free algebra.
    \end{proof}

    Due to close connections presented above of initial algebras and free
    algebras, from which algebraic effects are arising, we could not spare
    a few words a about induction and coinduction, where the latter may give
    us more intuition about coalgebraic effects.

    \textit{Induction} is a way of constructing new structures. \textit{Recursion} is
    a way of folding inductively defined structure in a terminating way. Recursive
    functions (not to be confused with recursive computability class), should
    shrink the argument in each call, meaning that it eventually ends up
    terminating in a base case.

    \textit{Coinduction} is literally a dual notion to induction. We \textit{observe}
    possibly infinite structures, by doing deconstruction. \textit{Corecursion}
    is a way of productively defining new, possibly enlarged structures. Due to
    infinity, the evaluation should be lazy, whereas in induction it may be eager.

@@

@@

    \subsection{Cooperations}\label{subsec:calc-cohandlers}

    Semantics for cooperations is brought from denotational one that we have
    introduced in Section~\ref{sec:duality}, meaning, we denote them
    as state passing transitions in a code, represented similarly as algebraic
    effects, through (co)free (co)models of algebraic theories along with their
    morphisms. Let's put the following cooperation:

@@

    Operationally, we embed transitions between coeffectful computations
    in a standard state monad $ S_{C}X \triangleq (C \Rightarrow X \times C) $.
    We achieve this by transforming AST of the source code, such that
    configurations are passed among cooperations. For a given coalgebraic theory
    $T$, handler of the state monad in Freak looks as follows:

@@

    \noindent
    into the following one (modulo $\mathVar{\alpha-conversion}$):

\vbox{\begin{verbatim}
    let Tconf <- observe TGet () in
    let (TnewConf, Tres) <- observe Coeffect (Tconf, p) in
    let _ <- observe TPut TnewConf in
    return Tres
\end{verbatim}
}

    \noindent
    where before executing a coeffect, we obtain current configuration,
    pass it to coeffect, put new config and then return the actual result.
    This is done behind the scenes and user does not know about state
    passing. Note that for readability we have done pattern matching on pairs,
    while currently in Freak it needs to be deconstructed using projections.

    \subsection{Cohandlers}

    Cohandlers needs to be aware of configuration passing, therefore,
    every cohandler accepts $ C \times P $ as a parameter where $C$ is current
    configuration and $P$ parameter passed to coeffect. In a similar way,
    value passed to resumption also needs to be a pair $ C \times X $ of new
    configuration and value returned down to computation.

    For a full currently implemented example of transformation, one may take a
    look at \verb!programs/transform/effCase! directory in source code.

@@

    State passing is defined as it was described in
    Subsection~\ref{subsec:calc-cohandlers}.
    From implementation standpoint, transformation is done by manipulating
    AST of the language to inject constructs handling state management.

@@

\begin{verbatim}
    AST.hs          - AST data structures
    CPSLists.hs     - Uncurried CPS translation
    CPSMonad.hs     - Definition of CPSMonad
    DValue.hs       - Evaluation result data structures
    EvalTarget.hs   - Evaluation of the target calculus
    Freak.hs        - API for the language
    Main.hs         - Main module running evaluator on given filename
    Parser.hs       - Parser and lexer
    TargetAST.hs    - AST for the target calculus
    Tests.hs        - Tests module
    Transform.hs    - Cohandlers transformation described in Calculus chapter
    Types.hs        - Common types definition
    programs/       - Exemplary programs covered in tests

@@

    We have showed derivation of cooperations, based on previous work, as well
    as presented a calculus combining algebraic effects with their dual notion.
    This resulted in a programming language which allows user to seemlessly
    operate on resources based on solid mathematical foundations. It can seen now,
    that (co)algebraic effects are an abstract concept that captures plenty of
    patterns in everyday programming, which already had ad-hoc and more
    restricted implementations in many existing languages, through constructs
    such as \verb!try, except, finally!, or by more low-level mechanisms
    like handling interrupts in operating systems.

    This relatively new concept in theory of programming languages yet needs to
    be researched, and we propose in the next section a few directions into
    which Freak could go.
